Distributional Regression Forests for Probabilistic Modeling and Forecasting

### 2018-02-10 pkg/disttree/inst/Psychoco2018/slides.Rnw
tried to reduce complexity in estimation and simulation sections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2008 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3447011d4cf715bb6885af7875d01db529eecda0

\documentclass[11pt,t,usepdftitle=false,aspectratio=43]{beamer}
%\usetheme[nototalframenumber, noslidenumber, url]{uibk}
\usetheme[nototalframenumber, noslidenumber]{uibk}

@@

\textbf{Distributional:} \\
\vspace{0.1 cm}
\hspace{0.5cm} Specify the complete probability distribution\\
\hspace{0.5cm} (including location, scale, and shape).

@@

{\bf Tree:}
\begin{itemize}
  \item Automatic detection of steps and abrupt changes. %(data driven)
  \item Capture non-linear and non-additive effects and interactions.
\end{itemize}

@@

{\bf Forest:}
\begin{itemize}
  \item Smoother effects.
  \item Stabilization and regularization of the model.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Global likelihood estimation}
\begin{itemize}
\item Specify distribution family with distribution function 
$$F(y) = \mathbb{P}_{\theta}(Y\leq y)$$ 
and log-likelihood function $$\ell(\theta; y).$$ 
The full distribution is then defined by the parameter vector $\theta \in \Theta$ representing location, scale, and shape.

\item Estimate $\hat{\bold{\theta}}$ via maximum likelihood 
$$
\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i).
$$
for a learning data set $\{y_i\}_{i = 1,\ldots,n}.$
%% equivalent to solving 
%% $$
%% \sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i) = \sum_{i=1}^n s(y_i, \hat{\theta}) = 0.
%% $$
\end{itemize}

@@

$$
\hat{\theta}(\bold{x}) = \max_{\theta \in \Theta} \sum_{i=1}^n w_i(\bold{x}) \cdot \ell(\theta; y_i).
$$

@@

How to calculate weights $\{w_i(\bold{x})\}_{i=1,\dots,n}$?\\

@@

Two strategies:
\begin{itemize}
\item Build a tree and let $w_i(\bold{x}) \in \{0,1\}$ indicate whether $\bold{x}$ is in the same subgroup as the $i$-th learning observation.
\item Build an ensemble of trees on resampled versions of the learning data and let $w_i(\bold{x}) \in  \mathbb{N}_0$ count the number of trees in which $\bold{x}$ is in the same subgroup as the $i$-th learning observation.
%\item Via a tree model: $w_i(\bold{x}) \in \{0,1\}$ indicates whether $\bold{x}$ is in the same terminal node as the $i$-th learning observation.
%\item Via a forest model: $w_i(\bold{x}) \in  \mathbb{N}_0$ counts the number of trees in which $\bold{x}$ is in the same terminal node as the $i$-th learning observation. 
\end{itemize}

@@

\subsection{Building distributional trees and forests}
\begin{frame}
\frametitle{Building distributional trees and forests}
{\bf Tree:}
\begin{enumerate}
\item Estimate $\hat{\theta}$ via maximum likelihood.
\item Test for associations or instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each partitioning variable $x_i$.
\item Split the sample along the partitioning variable with the strongest association or instability.
    Choose breakpoint with highest improvement in log-likelihood.
\item Repeat steps 1--3 recursively until some stopping criterion is met, yielding
    $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
\end{enumerate}

@@

{\bf Forest:} Ensemble of $T$ trees.
\begin{itemize}
\item Bootstrap or subsamples.
\item Random input variable sampling.
\end{itemize}
\end{frame}

@@

{\bf Estimator:}
$$
\hat{\theta}(\bold{x}) =  \max_{\theta \in \Theta} \sum_{i=1}^n w_i(\bold{x}) \cdot \ell(\theta; y_i)
$$

@@

\begin{frame}
\frametitle{Simulation}
{\bf Models}:

\smallskip

\hspace{0.55 em} disttree, distforest (100 trees), gamlss\\

@@

{\bf DGP}:

\smallskip

\begin{tabular}{ll}
Response:         & $y \sim \mathcal{N}(\mu(x), \sigma(x))$ with\\
                  & \qquad $\mu(x) =  10 \cdot \exp\left(-(4 \cdot x - 2)^{4}\right)$\\
                  & \qquad $\sigma(x) = 0.5 + 2 \cdot |x|$\\
Covariate:        & $x \sim  \mathcal{U}([-0.4,1])$\\
\# observations:  & 300 
\end{tabular}
\end{frame}

@@

{\bf Models}:

\footnotesize
\hspace{0.55 em} disttree, distforest (100 trees), gamlss\\

\normalsize
{\bf DGP}:

\smallskip

\footnotesize
\begin{tabular}{ll}
Response:         & $y \sim \mathcal{N}(\mu(x), \sigma(x))$ with\\
                 & \qquad $\mu(x) =  10 \cdot \exp\left(-(4 \cdot x - 2)^{2 \cdot \kappa}\right)$\\
                 & \qquad $\sigma(x) = 0.5 + 2 \cdot |x|$\\
                 & (increasing $\kappa$ leads to steeper transitions)\\
Covariate:        & $x \sim  \mathcal{U}([-0.4,1])$\\
\# observations:  & 1000 
\end{tabular}

\normalsize
{\bf Evaluation} over increasing $\kappa$:

\smallskip

\footnotesize
\begin{tabular}{ll}
Interval for $\kappa$:     & $[1, 71]$ \\
Stepsize for $\kappa$:     & 7 \\
\# replications per step:  & 150 
\end{tabular}

@@

\medskip
{\bf Based on:}
\begin{itemize}
\item Recommendations by specialist literature.
\item Expert subject matter knowledge.
\item Experiences from other modeling situations.
\item Common simplifications/generalizations/assumptions.
\end{itemize}

\bigskip
{\bf Examples:}
\begin{itemize}
\item Survival times: Weibull distribution, exponential distribution.
\item Wages, precipitation: censored normal distribution.
\item Temperature: normal distribution.
\item Count of events/accidents (insurance): Poisson distribution.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Alternative: Transformation models}
{\bf Advantages:}
\begin{itemize}
\item Does not require specification of distribution family.
\item More flexible framework.
\end{itemize}

@@

{\bf Distribution function:}
$$
F(y) = \Phi(\bold{a}_{Bs,d}(y)^\top \theta) 
$$

\begin{itemize} 
\item $\bold{a}_{Bs,d}(y)^\top \theta$ is a smooth, monotone Bernstein polynomial of degree $d$.
\item $d=1$ corresponds to $\mathcal{N}(\mu,\sigma^2)$.
\item $d=5$ is surprisingly flexible.
\end{itemize}

@@

%\newgeometry{left=0.2cm, right=0.2cm, top = 0.5cm, bottom=0.5cm}
\begin{frame}
\frametitle{Alternative: Transformation model}

@@

\end{frame}
%\restoregeometry

@@

\begin{tabular}{ll}
\code{distfit}    & Distributional fits (ML, \code{gamlss.family}/custom \code{list}).\\
                  & No covariates. \\
\code{disttree}   & Distributional trees (\code{ctree}/\code{mob} + \code{distfit}).\\
                  & Covariates as partitioning variables. \\
\code{distforest} & Distributional forests (ensemble of \code{disttree}s).\\
                  & Covariates as partitioning variables.
\end{tabular}

@@

@@

Seibold H, Zeileis A, Hothorn T (2017).
  \dquote{Individual Treatment Effect Prediction for Amyotrophic Lateral Sclerosis Patients.}
  \textit{Statistical Methods in Medical Research}, 
  \textbf{12}(1), 45--63.
  \doi{10.1177/0962280217693034}

@@

%% \begin{frame}[fragile]
%% \frametitle{Simulation}
%% \begin{center}
%% \vspace*{-1cm}
%% \setkeys{Gin}{width=0.8\linewidth}
%% <<plotsim_RMSE_par, fig=TRUE, echo=FALSE>>=
%% plot_rmse(simlist, type = "par")
%% @
%% \end{center}
%% \end{frame}

%% \begin{frame}[fragile]
%% \frametitle{Simulation}
%% \begin{center}
%% \vspace*{-1cm}
%% \setkeys{Gin}{width=0.8\linewidth}
%% <<plotsim_RMSE_exp, fig=TRUE, echo=FALSE>>=
%% plot_rmse(simlist, type = "exp")
%% @
%% \end{center}
%% \end{frame}

>>>>

\documentclass[11pt,t,usepdftitle=false,aspectratio=43]{beamer}
%\usetheme[nototalframenumber, noslidenumber, url]{uibk}
\usetheme[nototalframenumber, noslidenumber, url]{uibk}

@@

\textbf{Distributional:}
\begin{itemize}
  \item Specify the complete probability distribution (including location, scale, and shape).
\end{itemize}

@@

\textbf{Tree:}
\begin{itemize}
  \item Automatic detection of steps and abrupt changes. %(data driven)
  \item Capture non-linear and non-additive effects and interactions.
\end{itemize}

@@

\textbf{Forest:}
\begin{itemize}
  \item Smoother effects.
  \item Stabilization and regularization of the model.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Global likelihood estimation}
\begin{itemize}

\item Specify a parametric distribution family $F(\cdot; \theta)$ with parameter vector
  $\theta \in \Theta$ capturing location, scale, shape.

\item Cumulative distribution function and log-likelihood:
  \begin{eqnarray*}
  F(y; \theta)    & = & \mathbb{P}_{\theta}(Y\leq y)\\
  \ell(\theta; y) & = & \log(f(y; \theta))
  \end{eqnarray*}

\item Estimate $\hat{\bold{\theta}}$ via maximum likelihood
  based on a learning sample $y_1,\dots, y_n$:
\[
\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
\]

%% equivalent to solving 
%% $$
%% \sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i) = \sum_{i=1}^n s(y_i, \hat{\theta}) = 0.
%% $$
\end{itemize}

@@

\textbf{Idea:} Covariates captured through adaptive weights.
\[
\hat{\theta}(\bold{x}) = \max_{\theta \in \Theta} \sum_{i=1}^n w_i(\bold{x}) \cdot \ell(\theta; y_i).
\]

@@

\textbf{Question:} How to choose weighting function $w_i(\bold{x})$?

@@

\textbf{Possible answers:} Based on learning sample $y_1, \dots, y_n$ and (possibly new) observation $\bold{x}$.
\begin{itemize}
\item \emph{Tree:} $w_i(\bold{x}) \in \{0,1\}$ indicates whether $\bold{x}$ and $y_i$ are classified into the same subgroup.
\item \emph{Forest:} $w_i(\bold{x}) \in  [0, 1]$ averages the weights for $\bold{x}$ and $y_i$ across trees.
\end{itemize}

@@

\subsection{Building distributional trees and forests}
\begin{frame}
\frametitle{Distributional trees and forests}
\textbf{Tree:}
\begin{enumerate}
\item Estimate $\hat{\theta}$ via maximum likelihood (without covariates).
\item Test for associations or instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each partitioning variable $x_i$.
\item Split the sample along the partitioning variable with the strongest association or instability.
    Choose breakpoint with highest improvement in log-likelihood.
\item Repeat steps 1--3 recursively until some stopping criterion is met, yielding
    $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
\end{enumerate}

@@

\textbf{Forest:} Ensemble of $T$ trees.
\begin{itemize}
\item Bootstrap or subsamples.
\item Random input variable sampling.
\end{itemize}
\end{frame}

@@

\textbf{Estimator:}
\[
\hat{\theta}(\bold{x}) =  \max_{\theta \in \Theta} \sum_{i=1}^n w_i(\bold{x}) \cdot \ell(\theta; y_i)
\]

@@

\textbf{Models:} \code{disttree}, \code{distforest} (100 trees), \code{gamlss}.

@@

\textbf{Data:}
\begin{eqnarray*}
  y         & \sim & \mathcal{N}(\mu(x), \sigma(x))\\
  x         & \sim & \mathcal{U}(-0.4, 1) \\
  \mu(x)    & =    & 10 \cdot \exp\left\{ -(4 \cdot x - 2)^{2 \cdot \kappa} \right\}\\
  \sigma(x) & =    & 0.5 + 2 \cdot |x|
\end{eqnarray*}

\textbf{Parameters:}
\begin{itemize}
  \item 1 replication: $n = 300$, $\kappa = 2$.
  \item 150 replications: $n = 1000$, $\kappa = 1, 8, 15, \dots, 71$.
\end{itemize}
\end{frame}

@@

@@

\medskip
\textbf{Based on:}
\begin{itemize}
\item Recommendations by specialist literature.
\item Expert subject matter knowledge.
\item Experiences from other modeling situations.
\item Common simplifications/generalizations/assumptions.
\end{itemize}

\bigskip
\textbf{Examples:}
\begin{itemize}
\item Survival times: Weibull distribution, exponential distribution.
\item Wages, precipitation: censored normal distribution.
\item Temperature: normal distribution.
\item Count of events/accidents (insurance): Poisson distribution.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Transformation models}

\textbf{Advantages:}
\begin{itemize}
\item Does not require specification of distribution family.
\item More flexible framework.
\end{itemize}

@@

\textbf{Distribution function:}
\[
F(y; \theta) = \Phi(\bold{a}_{Bs,d}(y)^\top \theta) 
\]
\begin{itemize} 
\item $\bold{a}_{Bs,d}(y)^\top \theta$ is a smooth, monotone Bernstein polynomial of degree $d$.
\item $d=1$ corresponds to $\mathcal{N}(\mu,\sigma^2)$.
\item $d=5$ is surprisingly flexible.
\end{itemize}

@@

\begin{frame}
\frametitle{Transformation models}

@@

@@

\begin{tabular}{ll}
\code{distfit}    & Distributional fit (ML, \code{gamlss.family}/custom \code{list}).\\
                  & No covariates. \\
\code{disttree}   & Distributional tree (\code{ctree}/\code{mob} + \code{distfit}).\\
                  & Covariates as partitioning variables. \\
\code{distforest} & Distributional forest (\code{disttree} ensemble).\\
                  & Covariates as partitioning variables.
\end{tabular}

@@

Schlosser L, Hothorn T, Stauffer R, Zeileis A (2018).
  \dquote{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.}
  \emph{Manuscript under preparation}.

@@

@@
